---
title: "Regression Basics"
author: "Robert Gentleman"
format:
  revealjs:
    theme: [simple, ./custom.scss]
    slide-number: true
---

```{r}
#| echo: false
#| warning: false
#| message: false


library(GGally)
library(ggplot2)


```
## Outline

- basic ideas behind statistical methods
- why regression is an important tool for understanding data
- how to fit and interpret some regression models

## Where do data come from?

- experiments where someone has performed a task, usually varying some of the treatments or exposures and measuring outputs

- random sampling of some reasonably well defined population with the intention of making valid statements about that population

- clinical trials are like experiments but we don't control how the patients are selected

## Basics of Statistical Methods

::: {.incremental}

- when data are generated with variability we must account for that variation in our interpretation
- suppose we want to compare the cost to rent an apartment in two different cites
- in city A we sample 100 renters and get the mean rent per month of $500
- in city B we sample 100 renters and get the mean rent per month of $800
- which city is most expensive to rent in?
- you can't tell without knowing how much variation there is

:::

## Basics of Statistical Methods

::: {.incremental}

- *Statistics is about coming to conclusions in the face of uncertainty*, Wild and Seber, 2000.

- given data that are measured with error we need to extract information and estimate uncertainty

- we use models to help better estimate quantities of interest when they are affected by other factors

- A challenging part of the analysis is to identify the appropriate way to estimate the variation in the data

- Failing to do these steps well means you can (and often do) get the wrong answer.

:::

## Case study: is height hereditary?

- To understand the concepts of correlation and simple regression, we actually use the dataset from which regression was born.

- The example is from genetics.

- Francis Galton studied the variation and heredity of human traits.

- Among many other traits, Galton collected and studied height data from families to try to understand heredity.



## Case study: is height hereditary?

- While doing this, he developed the concepts of correlation and regression.

- At the time this data was collected, our knowledge of genetics was limited compared to what we know today.

- A very specific question Galton tried to answer was: **how well can we predict a child's height based on the parents' height?**



## Case study: is height hereditary?


- We have access to Galton's family height data through the **HistData** package.

- This data contains heights on several dozen families: mothers, fathers, daughters, and sons.

- To imitate Galton's analysis, we will create a dataset with the heights of fathers and a randomly selected son of each family:

```{r, message=FALSE, warning=FALSE, cache = FALSE} 
library(tidyverse) 
library(HistData) 
set.seed(1983) 
galton_heights <- GaltonFamilies |> 
  filter(gender == "male") |> 
  group_by(family) |> 
  sample_n(1) |> 
  ungroup() |> 
  select(father, childHeight) |> 
  rename(son = childHeight) 
``` 

- Suppose we were asked to summarize the father and son data.

```{r, message=FALSE, warning=FALSE} 

galton_heights |>  
  summarize(mean(father), sd(father), mean(son), sd(son)) 
``` 

## Case study: is height hereditary?

- This summary fails to describe an important characteristic of the data: 


```{r scatterplot-run, fig.align = "center", fig.height = 3, fig.width = 3, out.width="40%", echo=FALSE} 
galton_heights |> ggplot(aes(father, son)) +  
  geom_point(alpha = 0.5) 
``` 

## The correlation coefficient


- The correlation coefficient is always between -1 and 1.

- Pearson correlation measures a linear relationship between x and y

- the sample correlation coefficient is 1 or -1 only when x and y lie exactly on a straight line.

- The correlation, computed with the function `cor`, between father and son's heights is about 0.5:

::: {.large-output}

```{r}
#| echo: true
  
galton_heights |> summarize(r = cor(father, son)) |> pull(r) 
``` 

:::

---


## The correlation coefficient

-  Correlations ranging from -0.9 to 0.99:

```{r what-correlation-looks-like, fig.align = "center", echo=FALSE} 
n <- 250 
cors <- c(-0.9,-0.5,0,0.5,0.9,0.99) 
sim_data <- lapply(cors,function(r) MASS::mvrnorm(n,c(0,0), matrix(c(1,r,r,1),2,2))) 
sim_data <- Reduce(rbind, sim_data) 
sim_data <- cbind( rep(cors, each = n), sim_data) 
colnames(sim_data) <- c("r","x","y") 
as.data.frame(sim_data) |>  
  ggplot(aes(x,y)) +  
  facet_wrap(~r) +  
  geom_point(alpha = 0.5) + 
  geom_vline(xintercept = 0,lty = 2) +  
  geom_hline(yintercept = 0,lty = 2)  
``` 

## Sample correlation is random

- Before we continue connecting correlation to regression, let's remind ourselves about random variability.

- In most applications, we observe data that includes random variation.

- For example, in many cases, we do not observe data for the entire population of interest, but rather for a random sample.

- As with the average and standard deviation, the *sample correlation* is the most commonly used estimate of the population correlation.


## Sample correlation is random

- while the true correlation in the population is fixed but unkownn

- the correlation we compute on the sample and use as a summary is a random variable.

- By way of illustration, let's assume that the `r nrow(galton_heights)` pairs of fathers and sons is our entire population.

- A less fortunate geneticist can only afford measurements from a random sample of 25 pairs.

- The sample correlation can be computed with:

```{r} 
#| echo: true
r <- sample_n(galton_heights, 25, replace = TRUE) |>  
  summarize(r = cor(father, son)) |> pull(r) 
``` 

- `r` is a random variable.

## Sample correlation is random

- We can run a Monte Carlo simulation to see its distribution:


:::: {.columns}

::: {.column width="50%"}

```{r}
#| echo: false
#| fig-align: center
#| fig-height: 3
#| fig-width: 4
#| out-width: 100%
B <- 1000 
N <- 25 
r <- replicate(B, { 
  sample_n(galton_heights, N, replace = TRUE) |>  
    summarize(r = cor(father, son)) |>  
    pull(r) 
}) 
hist(r, breaks = 20) 
abline(v = 0.433, col = "red", lwd = 3)

```

:::

::: {.column width="50%"}

```{r}
#| echo: false
#| fig-align: center
#| fig-height: 3
#| fig-width: 4
#| out-width: 100%

galton_heights |>
  ggplot(aes(father, son)) +
  geom_point(alpha = 0.5)
```

:::

::::



## Sample correlation is random

- We see that the observed mean of `r` is close to the population correlation of $0.433$

```{r} 
mean(r) 
``` 

- and that the standard error is large relative to the range of values `r` can take:

```{r} 
sd(r) 
``` 

- So, when interpreting correlations, remember that correlations derived from samples are estimates with uncertainty.


## Correlation is not always useful 

:::: {.columns}

::: {.column width="50%"}


```{r ascombe-quartet, echo=FALSE,fig.width=7, fig.height=5.5, out.width="100%", fig.align="center"} 
anscombe |> mutate(row = seq_len(n())) |> 
  gather(name, value, -row) |>  
  separate(name, c("axis", "group"), sep=1) |> 
  spread(axis, value) |> select(-row) |> 
  ggplot(aes(x,y)) + 
  facet_wrap(~group)  + 
  geom_smooth(method="lm", fill=NA, fullrange=TRUE) +  
  geom_point() 
#+  geom_point(bg="orange", color="red", cex=3, pch=21) 
``` 

:::

::: {.column width="50%"}

- The following four artificial datasets, referred to as Anscombe's quartet, famously illustrate this point.

- All these pairs have a correlation of 0.82:

- To help us understand when correlation is meaningful as a summary statistic, we return to the example of predicting a son's height using his father's height.

:::

::::

## An Important Observation

:::: {.columns}

::: {.column width="50%"}

```{r}
#| echo: false
#| fig-width: 4
#| fig-height: 3
#| out-width: 100%

galton_heights |>
  ggplot(aes(father, son)) +
  geom_point(alpha = 0.5)
```

:::

::: {.column width="50%"}

- notice that there is a strong relationship between the height of a father and his son
- we will get better estimates if we can use the height of the father 
- statistical models are how we do that
- and we will first explore an *idea* called conditional expectation

:::

::::


## Conditional expectations


- Suppose we are asked to guess the height of a randomly selected son and we don't know his father's height.

- Because the distribution of sons' heights is approximately normal, we know the average height, `r round(mean(galton_heights$son), 1)`, is the value with the highest proportion and would be the prediction with the highest chance of minimizing the error.

- But what if we are told that the father is taller than average, say 72 inches tall, do we still guess `r round(mean(galton_heights$son), 1)` for the son?

- The statistical notation for the conditional expectation is:
$$
\mbox{E}(Y \mid X = x) 
$$
- with $x$ representing the fixed value that defines that subset, for example 72 inches.
- this represents the average value of the variable named $Y$ when the variable named $X$ has value $x$


## Conditional expectations

:::: {.columns}

::: {.column width="50%"}


- we divide the data up into groups depending on the father's height

- for each group we compute a boxplot

- when we examine the boxplots we see that the son's heights really change depending on the father's height


:::

::: {.column width="50%"}

```{r boxplot-1-run, fig.align = "center", fig.height = 3, fig.width = 3, out.width="100%", echo=FALSE} 
galton_heights |> mutate(father_strata = factor(round(father))) |>  
  ggplot(aes(father_strata, son)) +  
  geom_boxplot() +  
  geom_point() 
``` 

:::

::::


## Conditional expectations

:::: {.columns}

::: {.column width="50%"}


- Here we plot the means of the son's height for a binned value of the father's height
- These would be reasonable estimates to use
- But they waste information and are less precise than they could be
- We introduce models as ways to be more specific about the form of the relationship between the y's and the x's **and** to improve the precision of our estimates

:::

::: {.column width="50%"}

```{r conditional-averages-follow-line, fig.align = "center", echo=FALSE, fig.height = 3, fig.width = 3, out.width="100%"} 
galton_heights |>  
  mutate(father = round(father)) |>  
  group_by(father) |> 
  summarize(son_conditional_avg = mean(son)) |> 
  ggplot(aes(father, son_conditional_avg)) +  
  geom_point() 
``` 

:::

::::


## Regression

:::: {.columns}

::: {.column width="50%"}


```{r}
#| echo: false
#| fig-width: 6
#| fig-height: 5

set.seed(123)

x <- runif(20, 10, 20)
e <- rnorm(20, sd = 2)
y <- 3 + 2 * x + e

fit <- lm(y ~ x)
y_hat <- fitted(fit)

plot(
  x, y,
  type = "n",
  xlab = "",
  ylab = "",
  ylim = range(c(y, y_hat)),
  axes = FALSE
)

segments(
  x0 = x,
  y0 = y,
  x1 = x,
  y1 = y_hat,
  col = "red",
  lwd = 1
)

abline(
  fit,
  col = "cornflowerblue",
  lwd = 2
)

points(
  x, y,
  pch = 19,
  cex = 1.2
)

box()
```

:::

::: {.column width="50%"}

- the regression line is in blue
- the method used to estimate it is called least squares (LS)
- the LS solution minimizes the length of the red lines (they represent the *residuals*)

- we write this as  $y = \alpha + \beta \cdot x + \epsilon$

- and our regression line is: $E[Y|x] = \widehat{\alpha} + \widehat{\beta} \cdot x$

:::

::::

## The regression line

- we can show that the estimates of the intercept $\widehat{\alpha}$ and the slope $\widehat{\beta}$ are good in a variety of ways

- and we can use the estimated regression line to make predictions

- $E[Y| x_{new}] =  \widehat{\alpha} + \widehat{\beta} \cdot x_{new}$


## More general models

::::: {.columns}

:::: {.column width="50%"}


```{r}
#| echo: false
#| fig-width: 6
#| fig-height: 5

library(splines)

set.seed(123)

x <- runif(20, 10, 20)
e <- rnorm(20, sd = 2)

f <- function(x) {
  3 + 2 * x + 5 * sin(pi * (x - 10) / 10)
}

y <- f(x) + e

fit <- lm(y ~ ns(x, df = 4))
y_hat <- fitted(fit)

x_grid <- seq(10, 20, length.out = 200)
y_grid <- predict(fit, newdata = data.frame(x = x_grid))

plot(
  x, y,
  type = "n",
  xlab = "",
  ylab = "",
  axes = FALSE,
  ylim = range(c(y, y_grid))
)

box()

segments(
  x0 = x,
  y0 = y,
  x1 = x,
  y1 = y_hat,
  col = "red",
  lwd = 1
)

lines(
  x_grid,
  y_grid,
  col = "cornflowerblue",
  lwd = 2
)

points(
  x, y,
  pch = 19,
  cex = 1.2
)
```

::::

:::: {.column width="50%"}

- we want to choose a candidate that is smooth (usually) and where we try to minize the distance between the observed data and the smooth fit

- then we have:

::: {.highlight-equation}

$E[Y|x] = s(x) + \epsilon$
:::

::: {.highlight-equation}

$E[Y|x] = \widehat{s}(x)$

:::

- predictions are 

::: {.highlight-equation}

$E[Y|x_{new}] = \widehat{s}(x_{new})$


:::

::::

:::::

## How should we think about the changes in height

- when data are random we often don't want to use the units they were measured in for comparisons

- the reason is that we want to make the same conclusions whether we measure height in inches or in meters

- so when the data are approximately Normal$(\mu, \sigma)$ we measure distance in units that correspond to $\sigma$

- the change in y is typically not assessed in the units of y (inches in our example)

- it is more important to think of the change as a function of the variability in y and specifically the number of "SDs"

## Conditional expectations

```{r}
#| echo: false

conditional_avg <- galton_heights |>  
  filter(round(father) == 72) |> 
  summarize(avg = mean(son)) |>  
  pull(avg) 
#conditional_avg 
``` 


- A father that is 72 inches tall is about 1 SD taller than the mean height of fathers

- our prediction for the height of the son is: `r conditional_avg` 

- which is also taller than the average height of sons 

- **but** only `r round((conditional_avg - mean(galton_heights$son)) /sd(galton_heights$son),2)` standard deviations larger than the height of the average son.

- The sons of 72 inch fathers have *regressed* towards the average height.


## Regression improves precision

- Let's compare the two approaches to prediction that we have presented:

  1. Round fathers' heights to closest inch, stratify, and then take the average.

  2. Compute the regression line and use it to predict.

- We use a Monte Carlo simulation sampling $N=50$ families:

## Regression improves precision
:::: {.columns}

::: {.column width="50%"}


```{r} 
#| echo: true


B <- 1000 
N <- 50 
set.seed(1983) 
conditional_avg <- replicate(B, { 
  dat <- sample_n(galton_heights, N) 
  dat |> filter(round(father) == 72) |>  
    summarize(avg = mean(son)) |>  
    pull(avg) 
  }) 
regression_prediction <- replicate(B, { 
  dat <- sample_n(galton_heights, N) 
  mu_x <- mean(dat$father) 
  mu_y <- mean(dat$son) 
  s_x <- sd(dat$father) 
  s_y <- sd(dat$son) 
  r <- cor(dat$father, dat$son) 
  mu_y + r*(72 - mu_x)/s_x*s_y 
}) 
``` 

:::
::: {.column width="50%"}

- Although the expected value of these two random variables is about the same:

```{r} 
#| echo: true

mean(conditional_avg, na.rm = TRUE) 
mean(regression_prediction) 
``` 

- The standard error for the regression prediction is substantially smaller:

```{r} 
#| echo: true

sd(conditional_avg, na.rm = TRUE) 
sd(regression_prediction) 
``` 

:::

::::

## Regression improves precision

- The prediction obtained with the regression line is therefore much more stable than the prediction obtained using the conditional mean.

- There is an intuitive reason for this.

- The conditional average is computed on a relatively small subset: the fathers that are about 72 inches tall.

- In fact, in some of the permutations we have no data, which is why we used `na.rm=TRUE` in our simulation.

- The regression line is estimated using all the data.


## Multivariate Regression

- if one variable is good - then **two** variables must be better....


$$
Y = \alpha + \beta_1 X_1 + \beta_2 X_2 + \ldots + \beta_k X_k + \epsilon
$$ 
 where again we assume $\epsilon ~ N(0, \sigma^2)$
 
 - we can fit these models using least squares as well to get estimates of the model parameters and make predictions
 
 - as before we will need to engage in a substantial amount of model building and diagnostic assessment
 
## An Example: The Boston Housing Data

- data on housing prices as a market measure of the value people place on environmental quality. 
- assumes that a home's price reflects not only the house itself but also desirable neighborhood characteristics such as cleaner air, lower crime, better schools, and access to jobs. 
- they hoped that by modeling house prices while controlling for these other characteristics, they could estimate the economic value of improved air quality.

- we will return to this data in the lab

## Boston Housing Data
```{r}
#| label: colinear-1
#| message: false
#| echo: false
#| warning: false
library("MASS")
data("Boston")
#head(Boston)
```


```{r}
library(DT)
#| label: bostonanno
#| echo: false
bostonanno = read.table(textConnection("
varname|meaning
zn|proportion of residential land zoned for lots over 25,000 sq.ft
indus|proportion of non-retail business acres per town
chas|Charles River proximity (1 if tract bounds river; 0 otherwise)
nox|nitrogen oxides concentration (parts per 10 million)
rm|average number of rooms per dwelling
age|proportion of owner-occupied units built prior to 1940
dis|weighted mean of distances to five Boston employment centres
rad|index of accessibility to radial highways
tax|full-value property-tax rate per $10,000
ptratio|pupil-teacher ratio by town
black|$1000(b−0.63)^2$, where $b$ is the proportion of blacks
lstat|lower status of the population (percent)
medv|median value of owner-occupied homes in $1000s"), sep = "|", header = TRUE)
datatable(
  bostonanno,
  rownames = FALSE,
  options = list(
    dom = "t",
    paging = FALSE,
    searching = FALSE,
    info = FALSE,
    scrollX = TRUE,
    scrollY = "400px"
  )
)
#knitr::kable(bostonanno)
```

## Boston Housing Data

```{r}
#| echo: true
fit1 = lm(medv ~ dis + rm, data = Boston)
summary(fit1)
```


## Model diagnostics: Residual Plot

```{r}
#| echo: true

plot(fit1$fitted.values, fit1$residuals, xlab="Residuals", ylab="Fitted Values", pch=19, col="slategrey")
```

## Pairs Plot
:::: {.columns}

::: {.column width="60%"}


```{r}
#| fig-width: 8
#| fig-height: 8
#| out-width: 100%
GGally::ggpairs(
  Boston,
  columns = c("medv", "rm", "dis", "nox", "crim"),
  lower = list(
    continuous = wrap("points", alpha = 0.4, size = 0.7)
  ),
  upper = list(
    continuous = wrap("cor", size = 4)
  ),
  diag = list(
    continuous = wrap("densityDiag", alpha = 0.5)
  ),
  progress = FALSE
) +
  theme_bw()
```

:::
::: {.column width="40%"}

- notice the very odd behavior of some of the features
- there are outliers, limits of detection, non-linearty
- and potential interactions
- this is a challenge to fit and interpret

:::

::::


## Multivariate Regression: Interactions

- sometimes the effect of $X_i$ depends on the value (level) of $X_j$

- E.g. Smoking and asbestos each increase the risk of lung cancer, but together they produce a much greater increase than either exposure alone.

- if we believe there might be an interaction then we fit it by using special syntax, $X_i * X_j$

## Interactions

```{r}
#| label: ii2
#| echo: true
fit1I = lm(medv ~ dis + rm + lstat + rm*lstat, data = Boston)
summary(fit1I)

```


## Boston Housing Data

- We do two linear regressions with $y$ the house value, and two predictors (`dis` and `rm`), or three (in addition, `nox`).

- what changes?

```{r}
#| label: colinear-2
#| echo: true
#fit1 = lm(medv ~ dis + rm, data = Boston)
coef(fit1)
fit2 = lm(medv ~ dis + rm + nox, data = Boston)
coef(fit2)
```

## Collinearity

```{r}
fit3 = lm(medv ~ dis + rm + rad + tax, data = Boston)
summary(fit3)

```

## Collinearity
- RAD – TAX have a correlation of about 0.9

```{r}
fit4 = lm(medv ~ dis + rm + rad, data = Boston)
summary(fit4)
```

## Simpson's Paradox

- one of the reasons that we need to be careful with multivariate models is that the potential covariates have very particular effects


## Simpson's Paradox

:::: {.columns}

::: {.column width="50%"}


```{r simpsons-paradox, echo=FALSE}
N <- 100
Sigma <- matrix(c(1,0.75,0.75, 1), 2, 2)*1.5
means <- list(c(x = 11, y = 3), 
              c(x = 9, y = 5), 
              c(x = 7, y = 7), 
              c(x = 5, y = 9), 
              c(x = 3, y = 11))
dat <- lapply(means, function(mu){
  res <- MASS::mvrnorm(N, mu, Sigma)
  colnames(res) <- c("x", "y")
  res
})

dat <- do.call(rbind, dat) |> 
  as_tibble() |>
  mutate(z = as.character(rep(seq_along(means), each = N)))

dat |> ggplot(aes(x, y)) + geom_point(alpha = 0.5) +
  ggtitle(paste("Correlation = ", round(cor(dat$x, dat$y), 2))) +
theme(
    plot.title = element_text(size = 20)
  )
```

You can see that $X$ and $Y$ are negatively correlated. However, once we stratify by $Z$ (shown in different colors below), another pattern emerges:

:::
::: {.column width="50%"}

```{r simpsons-paradox-explained, echo=FALSE}
means <- do.call(rbind, means) |> 
  as_tibble() |>
  mutate(z = as.character(seq_along(means)))
  
corrs <- dat |> group_by(z) |> summarize(cor = cor(x, y)) |> pull(cor)

dat |> ggplot(aes(x, y, color = z)) + 
  geom_point(show.legend = FALSE, alpha = 0.5) +
  ggtitle(paste("Correlations =",  paste(signif(corrs,2), collapse = " "))) +
  theme(plot.title = element_text(size = 20)) +
  annotate("text", x = means$x, y = means$y, label = paste("z =", means$z), cex = 5)  
```

It is really $Z$ that is negatively correlated with $X$. If we stratify by $Z$, the $X$ and $Y$ are actually positively correlated, as seen in the plot above.


:::
::::

## Confounding

- When we are unable to randomly assign each individual to a treatment or control groups,
confounding becomes particularly prevalent. 

- Confounding occurs when a third variable is associated with both the exposure and the outcome, creating a misleading or distorted estimate of their relationship.

- A confounder is not part of the causal pathway and should be addressed through study design or analysis, such as randomization, restriction, matching, stratification, or regression adjustment.

- For instance, consider estimating the effect of eating fast foods on life expectancy using data collected from a random sample of people in New York City. 

- Fast food consumers are more likely to be smokers, drinkers, and have
lower incomes. 

- Consequently, a naive regression model may lead to an overestimate of the
negative health effects of fast food. 

## Collider Bias

- when the selection process introduces a correlation when there is not one in the population

- Mary wants to find a boyfriend.  But she has high standards and he must be either good looking or he must be smart

- She then notices that the smart boys are not that good looking and the good looking boys are not that smart.

- this happens in many situations - especially for example when studying diseases in hospitals

- being in hospital for something means that you probably have that disease, but not others